### 2020-03-04 main_text/main_text.tex
NB main text edits

	% Title must be 250 characters or less.
	\begin{flushleft}
		{\Large
			\textbf\newline{Theoretical investigation of a genetic switch for metabolic adaptation} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
		}
		\newline
		% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
		\\
		Kathrin S. Laxhuber\textsuperscript{1},
		Muir J. Morrison\textsuperscript{2},
		Griffin Chure\textsuperscript{3},
		Rob Phillips\textsuperscript{2,3$\ast$}
		\\
		\bigskip
		\textbf{1} Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland
		\\
		\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
		\\
		\textbf{3} Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
		\\
		\bigskip

@@

	It is quite conceivable that this auto-activation process is common to many
	substances that a cell would want to consume. Such a switch enables the
	cell to sense and respond to its environment: if the substrate enters the
	cell, it activates the production of membrane transporters. The cell then
	starts accumulating the substrate, thereby ``testing'' the substrate's
	presence in the extracellular environment. If there is enough, the
	expression stabilizes at an ``on'' state and the cell has, in a short-term
	sense, adapted. When there is not enough substrate, the operon, which
	often encodes for a whole set of enzymes for this one metabolite, switches
	``off'' again. Such a mechanism could be involved in various cases of
	short-term adaptation.

@@

	Experimentally, we found that the expression level of \emph{xapAB} among
	cells is bimodal and that the system seems to be
	bistable (see next section). We aim to understand which of the circuit's
	features are necessary for bistability and investigate its behavior in
	different parameter regimes. After presenting some experimental background on the \emph{xap} genes, we discuss the
	details of our model. Next, we estimate the free parameters and then present
	the observations we made through phase diagrams, followed by the results
	from stochastic simulations.

@@

	As mentioned in
	the introduction, there are two binding sites for XapR on the
	\emph{xapAB} promoter: one directly at the polymerase binding site
	and one further upstream.
	Working in the \textDelta\emph{xapAB} background,
	we measured the expression level of our reporter when driven by two
	constructs: the wild-type \emph{xapAB} promoter
	(Fig~\subref{fig1:data}{B}, left panel), and the \emph{xapAB}
	promoter with the upstream binding site removed
	(Fig~\subref{fig1:data}{B}, right panel).
	Clearly, removing one XapR binding site dramatically reduced the
	responsiveness of the promoter to xanthosine (though not apparent
	from Fig~\subref{fig1:data}{B}, this response remains detectable).

@@

	\section*{Model}
	\subsection*{Step by step modeling of the system}
	In this section, we present our model of the \emph{xapAB} genetic circuit.
	Fig~\ref{fig2:model} shows an overview of this model.
	The qualitative picture of the circuit switching its state is as follows:
	\begin{itemize}
		\item In the initial absense of XapB, small amounts of xanthosine permeate
				into and out of the cell (discussed in more detail below).
		\item The presence of xanthosine shifts XapR's equilibrium from inactive to active.
		\item Active XapR binds to the \emph{xapAB} promoter, leading to transcription.
		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
		\item Production of more XapA and XapB is balanced by degradation or dilution through cell divisions.
	\end{itemize}
	Because xanthosine activates the transcription factor XapR, we have positive
	and negative feedback loops due to XapB and XapA, respectively. The remainder
	of this subsection discusses each of the above steps in detail, leading us
	to a set of two coupled ODEs. More in-depth explanations can be found
	in~\nameref{S1_Text}.

@@

	\paragraph*{Transcription.}
	Transcription and translation of the \emph{xapAB} gene, regulated by the
	induced XapR, produce the two proteins XapA and XapB. We start with
	transcription and assume that the binding of XapR and polymerase to the
	promoter is at quasi-equilibrium. The polymerase binding is modeled as
	independent of that of XapR, and all influence of the activator is pushed
	into the transcription rate. Furthermore, the binding energy of XapR to each
	of its two sites is assumed to be the same. A discussion of these
	simplifications can be found in~\nameref{S1_Text}.

@@

	\subsection*{Nondimensionalization}
	We have now formulated the behavior of the system in terms of the rate
	equations for mRNA, protein, and xanthosine. These equations can be
	nondimensionalized, which reduces the dimension of parameter space. We
	measure time in units of $\gamma_{\n{p}}^{-1}$ and concentrations in units
	of $K_{\n{a}}$ (except XapR, where the equations make it more natural to use
	$K_{\n{XapR}}$). In Table~\ref{table1:nondim}, all the nondimensional
	parameters and their definitions are listed. Furthermore, we define
	$\n{[m]_a} \defeq \frac{\n{[m]}}{K_{\n{a}}}$, $\n{[p]_a} \defeq
	\frac{\n{[p]}}{K_{\n{a}}}$, $\n{[x]_a} \defeq \frac{\n{[x]}}{K_{\n{a}}}$,
	and $\tau \defeq \gamma_{\n{p}} t$. Using these definitions, the following
	equations are obtained
	\begin{alignat}{2}
	\dd{\mathrm{[m]_a}}{\tau} &= \ &&
	\rho_{\n{m}} 
	\frac{
		\mathrm{[XapR]_{R,A}}^2 
		e^{- \Delta \epsilon_{\text{coop}}}
	}{
		1 + 
		2 \mathrm{[XapR]_{R,A}} +
		\mathrm{[XapR]_{R,A}}^2 e^{- \Delta \epsilon_{\text{coop}}}
	}
	- \gamma_{\n{mp}} \mathrm{[m]_a}
	\\
	\dd{\mathrm{[p]_a}}{\tau} &= \ &&
	\rho_{\n{p}} \mathrm{[m]_a}
	- \mathrm{[p]_a}
	\\
	\dd{\mathrm{[x]_a}}{\tau} &= \ && \left(k_{\n{\beta,i}} \frac{\n{[c]_a}}{K_{\n{\beta,i}} + \n{[c]_a}} - k_{\n{\beta,e}} \frac{\n{[x]_a}}{K_{\n{\beta,e}} + \n{[x]_a}} - k_{\n{\alpha}} \frac{\n{[x]_a}}{1 + \n{[x]_a}}\right) \mathrm{[p]_a} \\ & && + k_{\n{\eta}} \left(\mathrm{[c]_a} - \xi \mathrm{[x]_a}\right)
	\nonumber \\
	\text{with} & && \raggedleft [\n{XapR}]_{\mathrm{R,\n{A}}} = [\n{XapR}]_{\mathrm{R}} \frac{\left(1 + \frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}}\right)^2}{\left(1 + \frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}}\right)^2 + e^{\Delta \epsilon_{\n{x}}} \left(1+\frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}} \frac{1}{K_{\n{IA}}}\right)^2} \nonumber
	\end{alignat}

@@

	As already mentioned, we are working with one specific set of parameters
	here and we will now explain this choice of values. Firstly, they were
	picked roughly in the middle of the range that was estimated beforehand for this parameter (see Table~\ref{table1:nondim} and~\nameref{S1_Text}). Secondly, we
	chose parameters that allow clear bistability in the phase portraits as well as in the
	stochastic simulations (see later), which, of course, is not the case for any possible choice of parameters. Thirdly, by the corresponding choice of
	parameters it was ensured that the mRNA number per cell at the
	``switch-point'' is around 1: this is large enough to enable the system to
	clearly resolve the two stable fixed points (as we will see from the
	stochastic simulations later on), but is low enough to lead to mean mRNA
	numbers that are very reasonable (see~\cite{Milo2016} for the average mRNA
	numbers in bacterial cells). The protein and xanthosine
	concentrations followed from this, but with some variation in the parameters
	they could still be tuned to a certain extent.

@@

	\subsection*{Stochastic simulations}
	Stochastic simulations of the full 3-dimensional system of mRNA, protein,
	and xanthosine were run for comparison with the deterministic results.
	In~\nameref{S1_Text}, we present the underlying chemical master equation of
	the system. Because of the two different fixed points at low and high
	expression, the protein copy numbers in the problem vary from less than
	five to several thousand. Even worse, xanthosine copy numbers may range
	as high as $10^7$ at the high expression fixed point. For such large
	copy numbers, the number of reaction firings that must be simulated with
	Gillespie's classical algorithm leads to an impractical computational
	cost. This would make Gillespie's \texttau-leap algorithm ideal for the high
	expression state. On the other hand, \texttau-leaping cannot be used for
	the small protein copy numbers in the low expression state, or the mRNA
	copy number which remains of order ten or less in both states. For
	these reasons, we chose to work with the algorithm described
	in~\cite{Cao2006}, a hybrid form between Gillespie's classical and his
	\texttau-leap algorithm. We gratefully worked with the Python implementation
	of this algorithm in \emph{StochPy}, version 2.3~\cite{Maarleveld2015}.

@@

		\bibitem{Hammer-Jespersen1980}
		Hammer-Jespersen K, Buxton RS, Hansen TD.
		\newblock A second purine nucleoside phosphorylase in \emph{Escherichia coli}
		K-12.
		\newblock Molecular and General Genetics MGG. 1980;179(2):341--348.

		\bibitem{Seeger1995}
		Seeger C, Poulsen C, Dandanell G.
		\newblock Identification and characterization of genes (xapA, xapB, and xapR)
		involved in xanthosine catabolism in \emph{Escherichia coli}.
		\newblock Journal of bacteriology. 1995;177(7559336):5506--5516.

		\bibitem{Joergensen1999}
		Jorgensen C, Dandanell G.
		\newblock Isolation and characterization of mutations in the \emph{Escherichia
			coli} regulatory protein XapR.
		\newblock Journal of bacteriology. 1999;181(10400599):4397--4403.

		\bibitem{Norholm2001}
		Norholm MH, Dandanell G.
		\newblock Specificity and topology of the \emph{Escherichia coli} xanthosine
		permease, a representative of the NHS subfamily of the major facilitator
		superfamily.
		\newblock Journal of bacteriology. 2001;183(11466294):4900--4904.

@@

		\bibitem{Milo2016}
		Milo R, Phillips R.
		\newblock In: Cell biology by the numbers. Garland Science. 2016; 217.

@@

>>>>

	% Title must be 250 characters or less.
	\begin{flushleft}
		{\Large
			\textbf\newline{Theoretical investigation of a genetic switch for metabolic adaptation} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
		}
		\newline
		% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
		\\
		Kathrin S. Laxhuber\textsuperscript{1},
		Muir J. Morrison\textsuperscript{2},
		Griffin Chure\textsuperscript{3},
		Nathan M. Belliveau\textsuperscript{4},
		Charlotte Strandkvist\textsuperscript{5},
		Kyle L. Naughton\textsuperscript{6},
		Rob Phillips\textsuperscript{2,3$\ast$}
		\\
		\bigskip
		\textbf{1} Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland
		\\
		\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
		\\
		\textbf{3} Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
		\\
		\textbf{4} Howard Hughes Medical Institute, University of Washington, Seattle, WA, USA
		\\
		\textbf{5} Department of Systems Biology, Harvard Medical School, Boston, MA, USA
		\\
		\textbf{6} Department of Physics and Astronomy, University of Southern California, Los Angeles, CA, USA
		\\
		\bigskip

@@

	It is quite conceivable that this auto-activation process is common to many
	substances that a cell would want to consume. Such a switch enables the
	cell to sense and respond to its environment: if the substrate enters the
	cell, it activates the production of membrane transporters. The cell then
	starts accumulating the substrate, thereby ``testing'' the substrate's
	presence in the extracellular environment. If there is enough, the
	expression stabilizes at an ``on'' state and the cell has, in a short-term
	sense, adapted. When there is not enough substrate, the operon, which
	often encodes for a whole set of enzymes for this one metabolite, switches
	``off'' again. Such a mechanism could plausibly be involved in various cases of
	short-term adaptation such as, but not limited to, the \emph{lacZYA}, \emph{araFGH},
	and \emph{xylFGH} examples mentioned above.

@@

	Experimentally, we found that the expression level of \emph{xapAB} among
	cells is bimodal and that the system seems to be
	bistable (see next section). We aim to understand which of the circuit's
	features are necessary for bistability and investigate its behavior in
	different parameter regimes. After presenting some experimental background on the \emph{xap} genes, we discuss the
	details of our model. Lastly, we estimate the free parameters and then present
	the observations we made through phase diagrams, followed by the results
	from stochastic simulations.

@@

	As mentioned in
	the introduction, there are two binding sites for XapR on the
	\emph{xapAB} promoter: one directly at the polymerase binding site
	and one further upstream.
	Working in the \textDelta\emph{xapAB} background,
	we measured the expression level of our reporter when driven by two
	constructs: the wild-type \emph{xapAB} promoter
	(Fig~\subref{fig1:data}{B}, left panel), and the \emph{xapAB}
	promoter with the upstream binding site removed
	(Fig~\subref{fig1:data}{B}, right panel).
	Clearly, removing one XapR binding site dramatically reduced the
	responsiveness of the promoter to xanthosine (though not apparent
	from Fig~\subref{fig1:data}{B}, this response remains detectable
	in other measurements covering a smaller dynamic range).

@@

	\section*{Model}
	\subsection*{Step by step modeling of the system}
	In this section, we present our model of the \emph{xapABR} genetic circuit.
	Fig~\ref{fig2:model} shows an overview of this model.
	The qualitative picture of the circuit switching its state is as follows:
	\begin{itemize}
		\item In the initial absense of XapB, small amounts of xanthosine permeate
				into and out of the cell (discussed in more detail below).
		\item The presence of xanthosine shifts XapR's equilibrium from inactive to active.
		\item Active XapR binds to the \emph{xapAB} promoter, increasing transcription.
		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
		\item Production of more XapA and XapB is balanced by degradation or dilution through cell divisions.
	\end{itemize}
	Because xanthosine activates the transcription factor XapR, we have positive
	and negative feedback loops due to XapB and XapA, respectively. The remainder
	of this subsection discusses each of the above steps in detail, leading us
	to a set of two coupled ODEs. More in-depth explanations can be found
	in~\nameref{S1_Text}.

@@

	\paragraph*{Transcription.}
	Transcription and translation of the \emph{xapAB} operon, regulated by the
	induced XapR, produce the two proteins XapA and XapB. We start with
	transcription and assume that the binding of XapR and polymerase to the
	promoter is at quasi-equilibrium. The polymerase binding is modeled as
	independent of that of XapR, and all influence of the activator is pushed
	into the transcription rate. Furthermore, the binding energy of XapR to each
	of its two sites is assumed to be the same. A discussion of these
	simplifications can be found in~\nameref{S1_Text}.

@@

	\subsection*{Nondimensionalization}
	We have now formulated the behavior of the system in terms of the rate
	equations for mRNA, protein, and xanthosine. These equations can be
	nondimensionalized, which reduces the dimension of parameter space. We
	measure time in units of $\gamma_{\n{p}}^{-1}$ and concentrations in units
	of $K_{\n{a}}$ (except XapR, where the equations make it more natural to use
	$K_{\n{XapR}}$). In Table~\ref{table1:nondim}, all the nondimensional
	parameters and their definitions are listed. Furthermore, we define
	$\n{[m]_a} \defeq \frac{\n{[m]}}{K_{\n{a}}}$, $\n{[p]_a} \defeq
	\frac{\n{[p]}}{K_{\n{a}}}$, $\n{[x]_a} \defeq \frac{\n{[x]}}{K_{\n{a}}}$,
	$\n{[c]_a} \defeq \frac{\n{[c]}}{K_{\n{a}}}$,
	and $\tau \defeq \gamma_{\n{p}} t$. Using these definitions, the following
	equations are obtained
	\begin{alignat}{2}
	\dd{\mathrm{[m]_a}}{\tau} &= \ &&
	\rho_{\n{m}} 
	\frac{
		\mathrm{[XapR]_{R,A}}^2 
		e^{- \Delta \epsilon_{\text{coop}}}
	}{
		1 + 
		2 \mathrm{[XapR]_{R,A}} +
		\mathrm{[XapR]_{R,A}}^2 e^{- \Delta \epsilon_{\text{coop}}}
	}
	- \gamma_{\n{mp}} \mathrm{[m]_a}
	\\
	\dd{\mathrm{[p]_a}}{\tau} &= \ &&
	\rho_{\n{p}} \mathrm{[m]_a}
	- \mathrm{[p]_a}
	\\
	\dd{\mathrm{[x]_a}}{\tau} &= \ && \left(k_{\n{\beta,i}} \frac{\n{[c]_a}}{K_{\n{\beta,i}} + \n{[c]_a}} - k_{\n{\beta,e}} \frac{\n{[x]_a}}{K_{\n{\beta,e}} + \n{[x]_a}} - k_{\n{\alpha}} \frac{\n{[x]_a}}{1 + \n{[x]_a}}\right) \mathrm{[p]_a} \\ & && + k_{\n{\eta}} \left(\mathrm{[c]_a} - \xi \mathrm{[x]_a}\right)
	\nonumber \\
	\text{with} & && \raggedleft [\n{XapR}]_{\mathrm{R,\n{A}}} = [\n{XapR}]_{\mathrm{R}} \frac{\left(1 + \frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}}\right)^2}{\left(1 + \frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}}\right)^2 + e^{\Delta \epsilon_{\n{x}}} \left(1+\frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}} \frac{1}{K_{\n{IA}}}\right)^2} \nonumber
	\end{alignat}

@@

	As already mentioned, we are working with one specific set of parameters
	here and we will now explain this choice of values. Firstly, they were
	picked roughly in the middle of the range that was estimated beforehand for this parameter (see Table~\ref{table1:nondim} and~\nameref{S1_Text}). Secondly, we
	chose parameters that allow clear bistability in the phase portraits as well as in the
	stochastic simulations (see later), which, of course, is not the case for any possible choice of parameters. Thirdly, by the corresponding choice of
	parameters it was ensured that the mRNA number per cell at the
	``switch-point'' is around 1: this is large enough to enable the system to
	clearly resolve the two stable fixed points (as we will see from the
	stochastic simulations later on), but is low enough to lead to mean mRNA
	numbers that are very reasonable (of order $10^{-2}$ to $10^{1}$
	transcripts per gene per cell in \textit{E.\ coli}~\cite{Milo2016}).
	The protein and xanthosine
	concentrations followed from this, but with some variation in the parameters
	they could still be tuned to a certain extent.

@@

	\subsection*{Stochastic simulations}
	Stochastic simulations of the full 3-dimensional system of mRNA, protein,
	and xanthosine were run for comparison with the deterministic results.
	In~\nameref{S1_Text}, we present the underlying chemical master equation of
	the system. Because of the two different fixed points at low and high
	expression, the protein copy numbers in the problem vary from less than
	five to several thousand. Even worse, xanthosine copy numbers may range
	as high as $10^7$ at the high expression fixed point. For such large
	copy numbers, the number of reaction firings that must be simulated with
	Gillespie's classical algorithm~\cite{Gillespie1976} leads to an impractical computational
	cost. This would make Gillespie's \texttau-leap algorithm~\cite{Gillespie2001} ideal for the high
	expression state. On the other hand, \texttau-leaping cannot be used for
	the small protein copy numbers in the low expression state, or the mRNA
	copy number which remains of order ten or less in both states. For
	these reasons, we chose to work with the algorithm described
	in~\cite{Cao2006}, a hybrid form between Gillespie's classical and his
	\texttau-leap algorithm. We gratefully worked with the Python implementation
	of this algorithm in \emph{StochPy}, version 2.3~\cite{Maarleveld2015}.

@@

		\bibitem{Hammer-Jespersen1980}
		Hammer-Jespersen K, Buxton RS, Hansen TD.
		\newblock A second purine nucleoside phosphorylase in \emph{Escherichia coli}
		K-12.
		\newblock Molecular and General Genetics. 1980;179(2):341--348.

		\bibitem{Seeger1995}
		Seeger C, Poulsen C, Dandanell G.
		\newblock Identification and characterization of genes (xapA, xapB, and xapR)
		involved in xanthosine catabolism in \emph{Escherichia coli}.
		\newblock Journal of Bacteriology. 1995;177(7559336):5506--5516.

		\bibitem{Joergensen1999}
		Jorgensen C, Dandanell G.
		\newblock Isolation and characterization of mutations in the \emph{Escherichia
			coli} regulatory protein XapR.
		\newblock Journal of Bacteriology. 1999;181(10400599):4397--4403.

		\bibitem{Norholm2001}
		Norholm MH, Dandanell G.
		\newblock Specificity and topology of the \emph{Escherichia coli} xanthosine
		permease, a representative of the NHS subfamily of the major facilitator
		superfamily.
		\newblock Journal of Bacteriology. 2001;183(11466294):4900--4904.

@@

		\bibitem{Milo2016}
		Milo R, Phillips R.
		\newblock In: Cell Biology by the Numbers. Garland Science. 2016; 217.

@@

		\bibitem{Gillespie1976}
		Gillespie, DT.
		\newblock A General Method for Numerically Simulating the Stochastic
		Time Evolution of Coupled Chemical Reactions.
		\newblock Journal of Computational Physics. 1976;22(4):403--434.

		\bibitem{Gillespie2001}
		Gillespie, DT.
		\newblock Approximate Accelerated Stochastic Simulation
		of Chemically Reacting Systems.
		\newblock J Chem Phys. 2001;115(4):1716--1733.
